---
title: "Lab 18: Mini Project: Investigating Pertussis Resurgence"
author: "Joshua Robinson"
format: pdf
toc: true
---

```{r}
library(datapasta)
library(ggplot2)
```

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r}
cdc <- data.frame(
  Year = c(2025,2024,2023,2022,2021,2020,2019,2018,2017,2016,2015,2014,2013,2012,2011,
           2010,2009,2008,2007,2006,2005,2004,2003,2002,2001,2000,1999,1998,1997,1996,
           1995,1994,1993,1992,1991,1990,1989,1988,1987,1986,1985,1984,1983,1982,1981,
           1980,1979,1978,1977,1976,1975,1974,1973,1972,1971,1970,1969,1968,1967,1966,
           1965,1964,1963,1962,1961,1960,1959,1958,1957,1956,1955,1954,1953,1952,1951,
           1950,1949,1948,1947,1946,1945,1944,1943,1942,1941,1940,1939,1938,1937,1936,
           1935,1934,1933,1932,1931,1930,1929,1928,1927,1926,1925,1924,1923,1922),
  Cases = c(28783,43321,7063,3044,2116,6124,18617,15609,18975,17972,20762,32971,28639,
            48277,18719,27550,16858,13278,10454,15632,25616,25827,11647,9771,7580,7867,
            7298,7405,6564,7796,5137,4617,6586,4083,2719,4570,4157,3450,2823,4195,3589,
            2276,2463,1895,1248,1730,1623,2063,2177,1010,1738,2402,1759,3287,3036,4249,
            3285,4810,9718,7717,6799,13005,17135,17749,11468,14809,40005,32148,28295,
            31732,62786,60886,37129,45030,68687,120718,69479,74715,156517,109860,13379,
            109873,191890,191383,222202,183866,103188,227319,214652,147237,180518,2652,
            179135,215343,172559,166914,197371,161799,181411,202210,152003,165418,1641,107473)
)
```

```{r}
ggplot(cdc, aes(x = Year, y = Cases)) +
  geom_line(color = "blue", linewidth = 0.8) +
  geom_point(color = "blue", size = 1) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Pertussis (Whooping Cough) Cases in the United States",
    x = "Year",
    y = "Number of Reported Cases",
    caption = "Source: CDC"
  ) +
  theme_minimal()
```

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc, aes(x = Year, y = Cases)) +
  geom_line(color = "blue", linewidth = 0.8) +
  geom_point(color = "blue", size = 1) +
  geom_vline(xintercept = 1946, linetype = "dashed", color = "green", linewidth = 0.8) +
  geom_vline(xintercept = 1996, linetype = "dashed", color = "red", linewidth = 0.8) +
  annotate("text", x = 1946, y = 250000, label = "wP vaccine\n(1946)", 
           hjust = -0.1, color = "darkgreen", size = 3) +
  annotate("text", x = 1996, y = 250000, label = "aP vaccine\n(1996)", 
           hjust = -0.1, color = "red", size = 3) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Pertussis (Whooping Cough) Cases in the United States",
    x = "Year",
    y = "Number of Reported Cases",
    caption = "Source: CDC"
  ) +
  theme_minimal()
```

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine in 1996, pertussis cases gradually rebounded with major spikes around 2010 and 2012. This is likely because the aP vaccine provides shorter lasting immunity than the older wP vaccine, allowing cases to rise as protection fades over time.

```{r}
library(jsonlite)
```

```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE) 
```

```{r}
head(subject, 3)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

```{r}
library(lubridate)
```

```{r}
today()
```

```{r}
today() - ymd("2000-01-01")
```

```{r}
time_length( today() - ymd("2000-01-01"),  "years")
```

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
subject$age <- today() - ymd(subject$year_of_birth)

tapply(time_length(subject$age, "years"), subject$infancy_vac, mean)

t.test(time_length(age, "years") ~ infancy_vac, data = subject)
```
> Q8. Determine the age of all individuals at time of boost?

```{r}
subject$age_at_boost <- time_length(ymd(subject$date_of_boost) - ymd(subject$year_of_birth), "years")
head(subject$age_at_boost)
```

> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

```{r}
ggplot(subject) +
  aes(time_length(age, "year"), fill = as.factor(infancy_vac)) +
  geom_histogram(show.legend = FALSE) +
  facet_wrap(vars(infancy_vac), nrow = 2) +
  xlab("Age in years")
```

```{r}
library(dplyr)
library(jsonlite)
```

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/specimen", simplifyVector = TRUE)
titer <- read_json("https://www.cmi-pb.org/api/plasma_ab_titer", simplifyVector = TRUE)
```

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
meta <- inner_join(specimen, subject, by = "subject_id")
dim(meta)
head(meta)
```

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(titer, meta, by = "specimen_id")
dim(abdata)
names(abdata)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

> Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI, antigen) +
  geom_boxplot() +
  xlim(0, 75) +
  facet_wrap(vars(visit), nrow = 2)
```

> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

FIM2/3, FHA, PRN, and PT show the clearest increases in IgG titer levels over time, particularly after the booster vaccination. These antigens are directly targeted by the pertussis vaccine, unlike unrelated antigens such as TT and DT. ALso controls like OVA.

```{r}
ggplot(abdata) +
  aes(x = MFI_normalised, y = antigen, col = infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit), nrow = 2) +
  xlim(0, 75) +
  theme_bw()
```

```{r}
igg %>% filter(visit != 8) %>%
ggplot() +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  xlim(0,75) +
  facet_wrap(vars(infancy_vac, visit), nrow=2)
```

> Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
filter(igg, antigen=="OVA") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

```{r}
filter(igg, antigen == "PT") %>%
  ggplot() +
  aes(MFI_normalised, antigen) +
  geom_boxplot(show.legend = TRUE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

> Q16. What do you notice about these two antigens time courses and the PT data in particular?

OVA shows no change across visits as expected, while PT rises dramatically after boosting with wP individuals showing higher IgG responses than aP individuals.

> Q17. Do you see any clear difference in aP vs. wP responses?

Yes, wP primed individuals consistently show higher PT antibody titers after boosting, suggesting the whole cell vaccine primes a stronger long term immune memory than the acellular vaccine.

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

```{r}
abdata.21 <- abdata %>% filter(dataset == "2020_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```


> Q18. Does this trend look similar for the 2020 dataset?

Yes. The 2020 dataset shows a similar trend: PT IgG levels peak around day 14 post boost then decline.

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"
rna <- read_json(url, simplifyVector = TRUE) 
```

```{r}
ssrna <- inner_join(rna, meta)
```

> Q19. Make a plot of the time course of gene expression for IGHG1 gene (i.e. a plot of visit vs. tpm).

```{r}
ggplot(ssrna) +
  aes(visit, tpm, group=subject_id) +
  geom_point() +
  geom_line(alpha=0.2)
```

> Q20.: What do you notice about the expression of this gene (i.e. when is it at it’s maximum level)?

GHG1 expression peaks dramatically at visit 4 before dropping back down to baseline levels.

> Q21. Does this pattern in time match the trend of antibody titer data? If not, why not?

Yes. This matches the antibody titer trend, which makes sense since IGHG1 encodes an IgG antibody protein. Increased gene expression precedes and drives the rise in detectable IgG antibody levels.

